### 2020-11-18 popl21/popl21-rebuttal.txt

Thank you for your reviews!

## Review A

### Corrections

The quoted sentence was part of our discussion of _future_ work. It was not written as motivation for this paper.

Livelits can generate expressions, not just values. This includes expressions that have non-trivial computational behavior. 
For example, Fig 3 demonstrates image transformations.

## Review B

### Comments

Although the five examples in the paper are non-trivial, we agree that additional screenshots might help readers build 
better intuitions for the expressive power of livelits. Consequently, we would be happy to include 4-6 additional 
screenshots of novel livelits in the form of an "example gallery" in our final revision. None of these introduce
new technical content into the paper.

Our prior work on Graphite included several empirical studies which are just as relevant to livelits.
The present paper can be read as a technical complement to that qualitative/empirical paper [Omar et al, 2012].

The Hazel livelits implementation as described in the paper was functional at the time of submission
(in the livelits branch on Github). It will be sent for artifact evaluation and released publicly with the 
next major Hazel release before POPL. The public will be able to play with livelits in their browser. We are 
working on a number of ``quality of life'' improvements on an ongoing basis, but these do not change the 
content or contributions of the paper.

## Review C

### Corrections

MPS and Barista (via Citrus) were discussed in detail in the paper. These systems, as well as the others that 
the review mentions, fail to achieve multiple of the design goals from Sec. 1 (discussed on line 921). We
will include missing citations in the final revision, thank you.

Livelits could be defined using other livelits when using the definition mechanism defined in Sec. 3. It is not the case that ``the definitions for livelits are textual'' in this setting. Indeed, related work on literal macros supports using literal macros in literal macro definitions [Omar et al, 2018] and the same mechanisms could be deployed here (once Hazel has a proper module / package management system).

We believe pure functional programming in languages like Haskell and Elm is practical. 
That said, it is not true that livelits are limited to the pure functional setting. 
The issue is simply that running a program with side effects
during editing might not be _desirable_ in some settings, but that choice is up to the programmer.
For example, live programming in Jupyter lab notebooks often involves side effects 
occurring in arbitrary order depending on user interactions and programmers by and large accept this. 
Livelits could be added to Jupyter-like systems such that they always use this ambient state.
The issue could also be avoided by using algebraic effects, so that destructive effects could be ``shimmed'' 
during development. These considerations are not livelit specific -- they are inherited from the work 
in [Omar et al., POPL 2019] and indeed live programming itself is fundamentally in tension with 
side effects. We will include a more substantial discussion of this point in the final revision.

We do discuss livelit persistence in Sec. 3.1. Only the model is serialized. In our case, we use a generic S-expression
textual syntax for Hazel expressions, but this is an incidental implementation detail (Hazel programs are not
intended to be edited using text editors). We do discuss a proof-of-concept text editor with support for 
livelits in Sec. 5.2.

### Comments

Pattern livelits would be a straightforward extension of our system. The necessary theory for pattern macros
was developed in [Omar et al, 2018].

Declaration livelits can be subsumed by module livelits, which can in turn be subsumed by expression livelits
in a language that has first class modules. This was also briefly discussed in [Omar et al, 2018].

While our formalism of expansion does build on our prior macro system formalism, we see this as a strength
of our paper. Our treatment of splices is distinct, as is our treatment of hygiene. Closure collection is entirely
new. The purpose of the formalism was to make the mechanism precise and reduce it to its essential form.

Hazel functions that use GUI libraries must ultimately execute by extraction to OCaml, so the fact that 
we implemented our GUI code directly in OCaml was merely a matter of expediency. The Hazel language is still
evolving (in large part due to contributors who are not involved with this paper) and there is not yet a 
full-scale GUI widgets library. We plan to "self host" in the coming months before POPL.

The interface for selecting between closures is the same as the interface demonstrated in [Omar et al, POPL 2019]. We will include a screenshot in the final revision.

The capture list in Fig. 4 is empty because generated expansions do not use any libraries or definition site values
(they are simple tuples of numbers in this case). See [Omar et al, ICFP 2018] for a discussion on explicit capture.

## Review D

### Corrections

The ipywidgets library only allows simple non-compositional widgets (no splices inside) and these widgets do not have access to the surrounding 
closures or state so there is no support for liveness. 
None of the central examples in our paper could be implemented as ipywidgets. We will, however, make sure to cite this work in the final revisions.

### Comments

See our response to Review B regarding user studies.

Typed holes are central in that they enable closure collection, which is what enables liveness.

The core calculus includes sums, products, and recursive types because these are minimal type theoretic constructs sufficient to achieve
self-hosting of syntax. Alternative encoding schemes could work just as well (e.g. Church encodings in System F) but are more unwieldy.
Self-hosting is necessary because the expand function returns an encoding of syntax, which needs to be decoded (Sec. 4.2.1).

Splices are simply expressions that appear inside (i.e. are notionally spliced into) livelit GUIs.

We will cite Lerner 2020, which has related motivations, thank you.

We will use a different connective for expanding and elaborating in the final version, thank you.

### 2020-11-18 popl21/case-studies.tex
moved popl21 submission into a directory for archival

\section{Case Studies}\label{sec:case-studies}
Let us begin by considering livelits from the client programmer's perspective by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.
These case studies and examples have been implemented\todo{make this not a lie}{}
in Hazel, a browser-based live programming environment for a pure typed functional language in the
ML family. We assume basic familiarity with ML.\todo{cite Elm tutorial for readers who are not familiar?}{}

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
\begin{figure}
    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
\caption{Grading Example}
\label{fig:grading}
\end{figure}

@@

Perhaps the most common end-user application in this domain is the spreadsheet.
A spreadsheet allows the instructor to record student grades using a natural tabular interface,
visualize this data in one of a finite number of styles, and perform basic computations on this data,
with results updated live, i.e. automatically after each cell has been edited.
However, these affordances are limited. It is difficult to package up common operations and workflows
into reusable libraries, interact with the data using domain-specific visualizations,
and perform operations that have not been anticipated by the designers of the spreadsheet system
(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission
to the university's grading system).
General-purpose programming languages equipped with suitable libraries for working with numeric data
are flexible enough to handle these scenarios, but users of these languages
sacrifice the ability to receive live feedback and directly manipulate the data and visualizations thereof.

Livelits are able to address this tension.
Fig.~\ref{fig:grading} shows a Hazel program where the user has:\todo{label each of these 1-5 in figure?}{}
\begin{enumerate}
\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell,
\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses,
\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
(for the sake of space, we plot dots rather than a full histogram)\todo{should we do a histogram?}{},
\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
\end{enumerate}

\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:\todo{change code once figure is final}{}

\begin{lstlisting}
let grades = Dataframe (
  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
  [
    ("Cyrus", [0.0, 0.0, 0.0, 0.0, 0.0]),
    ("Nick", [0.0, 0.0, 0.0, 0.0, 0.0]),
    ("David", [0.0, 0.0, 0.0, 0.0, 0.0])
  ]) in
let averages = compute_averages(grades) in
let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
format_for_university(assign_grades(averages, cutoffs))
\end{lstlisting}

The programmer can inspect this expansion in Hazel.
Ideally, however, it would not be necessary to inspect the expansion
(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions})
to reason about types and binding.
After all, one does not need to look inside
function bodies to reason about types and binding.
Instead, in the words of Reynolds\todo{cite}{},
``type structure is a syntactic discipline for maintaining levels of abstraction''.
The livelits mechanism supports reasoning abstractly about types and binding by
several means.

To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the two livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none]
livelit $dataframe at Dataframe { ... }
livelit $cutoffs(avgs: List(Float))
  at (.A Float, .B Float, .C Float, .D Float) { ... }
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe}
(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
and the expansion type of \li{\$cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
Hazel summarizes the typing information from the livelit definition for the programmer when the cursor is on the livelit's name,
just as it displays a function's type when its name is under the cursor (not shown).\todo{could show if Hannah et al are finished}{}

@@

\paragraph{Indeterminate Results}
We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
splices and parameters can be incomplete, i.e. they can contain holes, so
evaluation does not always result in a value \cite{HazelnutLive}.
In particular, when a hole appears in elimination position, e.g. in function position of a function application,
 the result is instead an indeterminate expression.
When a livelit requests an evaluation result, it must be able to handle indeterminate results.
For example, if the parameter to \li{\$cutoffs} were a hole or some other indeterminate expression\todo{small screenshot?}{},
then it would have degraded functionality
(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements\todo{implement this}{}).
We will return to how livelit implementations handle this situation in Sec.~\ref{sec:implementation}.
\todo{make sure we cover this}{}

\subsection{Case Study: Image Transformations}\label{sec:image-transformation}
Show example of an image transformation pipeline going through livelits

This might also benefit from Parameterization

Show multiple calls with different example images + closure selector UIs

Go into more detail about how evaluation works + fill-and-resume mechanics (and efficiency nod)

Talk about probes?

We assume that the user has selected at most one closure for each livelit in the unexpanded expression.
The corresponding environment is applied as a substitution before attempting to evaluate the requested expression.
To maintain the context independence principle, this request must take the form of a closed function
of the splices.
As discussed in Sec.~\ref{sec:live-evaluation}, evaluation can fail if the free variables in the splices
overlap with the free variables in selected closure.

\subsection{Additional Examples}\label{sec:additional-examples}
It would be nice to have a gallery-style figure and a brief overview of some other case studies
and how they exercise the novel features of the livelits mechanism. Maybe some statistics on how
many lines of code it took.

Ideas:
\begin{itemize}
  \item derivation trees like Joomy's system (\url{https://joom.github.io/proof-tree-builder/src/})
\end{itemize}

>>>>

\section{Case Studies}\label{sec:case-studies}
Let us begin by considering livelits from the client programmer's perspective by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
% We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
These case studies have been implemented
in Hazel, a browser-based live programming environment for a pure typed functional language in the
ML family. We assume basic familiarity with ML.

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
\begin{figure}
    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
\caption{Grading with Livelits}
\label{fig:grading}
\end{figure}

@@

Perhaps the most common end-user application in this domain is the spreadsheet.
A spreadsheet allows the instructor to record student grades using a natural tabular interface,
visualize this data in one of a finite number of plot styles, and perform basic computations on this data,
with results updated live, i.e. automatically after each cell has been edited.
However, these affordances are limited. It is difficult to package up common operations and workflows
into reusable libraries, interact with the data using domain-specific visualizations,
and perform operations that have not been anticipated by the designers of the spreadsheet system
(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission
to the university's grading system).
General-purpose programming languages equipped with suitable libraries for working with numeric data
are flexible enough to handle these scenarios, but users of these languages
sacrifice the ability to receive live feedback and directly manipulate data and visualizations thereof.

Livelits are able to address this tension.
Fig.~\ref{fig:grading} shows a Hazel program where the user has:
\begin{enumerate}
\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell,
\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses,
\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
(for the sake of space, we plot dots rather than a full histogram),
\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
\end{enumerate}

\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:

\begin{lstlisting}
let grades = Dataframe (
  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
  [
    ("Alice", [24. +. 36. +. 33., 92., 83.5, 87.5, 95., 88.]),
    ("Bob", [61., 64., 98., 89.5, 70., 85.]),
    ("Carol", [75., 81., 73., 71., 82., 79.]),
    (* ... *)
  ]) in
let averages = compute_averages grades weights in
let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
format_for_university (assign_grades averages cutoffs)
\end{lstlisting}

The programmer can inspect this expansion in Hazel.
Ideally, however, it would not be necessary to inspect the expansion
(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions})
to reason about types and binding.
After all, one does not need to look inside
function bodies to reason about types and binding.
Instead, in the words of \citet{DBLP:conf/ifip/Reynolds83},
``type structure is a syntactic discipline for maintaining levels of abstraction''.
The livelits mechanism supports reasoning abstractly about types and binding by
several means.

To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the two livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none]
livelit $dataframe at Dataframe { ... }
livelit $grade_cutoffs(avgs: List(Float))
  at (.A Float, .B Float, .C Float, .D Float) { ... }
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe}
(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
and the expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
Hazel summarizes the typing information from the livelit definition when the cursor is on the livelit's name,
just as it displays a function's type when its name is under the cursor (not shown).

@@

\paragraph{Indeterminate Results}
We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
splices and parameters can be incomplete, i.e. they can contain holes, so
evaluation does not always result in a value \cite{HazelnutLive}.
In particular, when a hole appears in elimination position, e.g. in function position of a function application,
 the result is instead an indeterminate expression.
When a livelit requests an evaluation result, it must be able to handle indeterminate results.
For example, if the parameter to \li{\$grade_cutoffs} were a hole or some other indeterminate expression,
then it would have degraded functionality
(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
We will return to how livelit implementations handle this situation in Sec.~\ref{sec:livelit-definitions}.

\subsection{Case Study: Image Transformations}\label{sec:image-transformation}

\begin{figure*}
  \begin{center}
    \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-1.png}
      \caption{}
    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{img-filter-2.png}
      \caption{}
    \end{subfigure}
  \end{center}
  \caption{Case Study: Image Transformation. The image shown is determined based on the selected closure.}
  \label{fig:img-transformation}
\end{figure*}

Our next case study is a simple image transformation livelit, \li{\$img_filter},
demonstrated in Fig.~\ref{fig:img-transformation}. This livelit takes 
a URL to an image as a parameter, and contains two splices of type \li{Int},
one to adjust the contrast, and the other to adjust the brightness.
In this example, we have filled those splices with a \li{\$slider}, but 
as above we could enter any expression of type \li{Int}.
The livelit shows a live preview of the transformed image.
The expansion generates the necessary calls to image processing functions, 
not shown.

The novelty in this example is that the livelit appears inside a function, 
\li{classic}, which takes an image URL as input. At the bottom 
of the figure, \li{classic} is mapped over a list (here, containing only 
two elements for the sake of presentation). Consequently, there are now 
two closures associated with the application of \li{\$img_filter} in the 
body of \li{classic}. Rather than disabling live evaluation in situations like 
this, Hazel instead allows the programmer to select between the closures when 
the cursor is on the livelit expression (via a simple sidebar toggle, not shown, which is shared 
with the general closure selection UI \cite{HazelnutLive}). 
The live feedback is based on the selected closure.
This makes it easy to see how the image filter being designed here will affect a
number of example images by quickly toggling between closures. The underlying expansion remains abstract, i.e. it refers to the image via the \li{url} variable.

% \subsection{Additional Examples}\label{sec:additional-examples}
% It would be nice to have a gallery-style figure and a brief overview of some other case studies
% and how they exercise the novel features of the livelits mechanism. Maybe some statistics on how
% many lines of code it took.

% Ideas:
% \begin{itemize}
%   \item derivation trees like Joomy's system (\url{https://joom.github.io/proof-tree-builder/src/})
% \end{itemize}
